What Hast Earth to do With Mars: An explanation of Urbit's IPC interfaces

### 2024-02-05 mss.tex
fixed neals comments

\section{Introduction}

In the realm of software, two worlds exist in stark contrast: 
Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
This article sets forth on an expedition to unravel the mysteries of this Martian software landscape, 
exploring how the intricate processes of terrestrial technology can establish a dialogue with the Martian code.

@@

@@

\subsection{Arvo: The Heart of Martian Computing}
Arvo, also known as Urbit OS, represents a paradigm shift in operating system design. Unlike traditional operating systems that operate on preemptive multitasking and complex event networks, Arvo is a purely functional operating system, characterized by its deterministic nature and compact size. The entire Urbit stack is about 30,000 lines of code, with Arvo itself being only around 1,000 lines. This small codebase is intentional, reflecting a philosophy that system administration complexity is directly proportional to code size.

Arvo's design philosophy positions it not just as an operating system but as a new frontier in the peer-to-peer internet space. Unlike traditional operating systems like Windows, macOS, or Linux, Arvo doesn't aim to replace them; instead, it offers a unique user experience akin to a web browser, yet it is a fully-fledged OS in its own right. This distinctive approach enables Arvo to function within a virtual machine, theoretically capable of running on bare metal.

@@

@@

\section{Background and Literature}

Exposite the relevant background and literature.

\begin{enumerate}
  \item  Prefer to \texttt{enumerate}.
  \item  It is easier to refer to than \texttt{itemize}.
\end{enumerate}

\noindent
Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.

\section{Urbit's Implementation}

Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.

\lstset{language=C}
\begin{lstlisting}
#include <stdio.h>

double compute(float a, float b) {
    return a * a + b * b;
}

int main() {
    float x = 2.5;
    float y = 3.7;

    double result = compute(x, y);

    printf("Result: %lf\n", result);

    return 0;
}
\end{lstlisting}

\noindent
That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.

\begin{figure}
  \begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
  def hello_world():
      print("Hello, world!")
  \end{lstlisting}
\end{figure}

Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)

\section{A Wild Ernest Appears}

You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.

@@

To summarize, why did you write it?  Why do we care?  What impact should it have on Urbit development?

Your bibliography is a separate BibTeX file.  We use the \texttt{plainnat} bibliography style.  You can use \texttt{natbib} citation commands like \texttt{\textbackslash citep\{wikipedia\}} for parenthesized references.  Use \texttt{\textbackslash citet\{wikipedia\}} for inline references.  You can also use \texttt{\textbackslash citeauthor\{wikipedia\}} for the principal author's name.

``You can use traditional TeX---or LaTeX---representations'' \citep{Varney1987}.

“Or you can use fancy quotes—and symbols.”

\bibliographystyle{plainnat}
\bibliography{ernest}
\end{document}

>>>>

\section{Introduction}
%Section 1: can do explicit callback to Moron Labs post (we should add that to the standard mss.bib)
% When did urbit backoff on the earth only calls mars? (eyre and iris) Urbit had to run on a VM
In the realm of software, two worlds exist in stark contrast: 
Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
This article sets forth on an expedition to unravel the mysteries of this Martian software landscape, 
exploring how the intricate processes of terrestrial technology can establish a dialogue with the Martian code.

@@

In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says ``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth. The latter would be quite impossible, since Earth code is non-Maxwellian. There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.'' This however has never been true for Urbit as it was implemented. Urbit has two vanes \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and curl, respectively). One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.

@@

\subsection{Arvo: The Heart of Martian Computing}
Arvo, also known as Urbit OS, represents a paradigm shift in operating system design. Unlike traditional operating systems that operate on preemptive multitasking and complex event networks, Arvo is a purely functional operating system, characterized by its deterministic nature and compact size. The entire Urbit stack is about 80,000 lines of code, with Arvo itself being only around 2,000 lines. This small codebase is intentional, reflecting a philosophy that system administration complexity is directly proportional to code size.

@@

The architecture of Vere is bifurcated into two distinct parts: the \texttt{king} and the \texttt{serf}. The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth. In contrast, the \texttt{serf} operates as the process running the Arvo virtual machine (VM). This division of responsibilities allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.

Communication between the \texttt{king} and \texttt{serf} is achieved through an Inter-Process Communication (IPC) port. This IPC mechanism facilitates a continuous dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.

A core opcode in Urbit's assembly language (\texttt{nock}) is the \texttt{hint} opcode (opcode 11) this allows Urbit to pass hints to Vere which produces effects on Earth. This bridges the Earth/Mars gap and allows for vanes to interact outside of Mars.

@@

\section{\%khan - The High-Level Thread Interface of Urbit}

Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem. It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad. This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.

Khan allows threads to be triggered from outside of Urbit. A thread is a monadic function that takes arguments and produces a result. It may perform input and output while running, so it is not a pure function. A thread's strength is that it can easily perform complex IO operations. It uses what's often called the IO monad (plus the exception monad) to provide a natural framework for IO. A thread's weakness is that it's impermanent and may fail unexpectedly. In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails. When code is upgraded, it's impossible to upgrade a running thread, so it fails.

Khan's role in Urbit can be likened to that of a control plane. Its primary function is to execute threads via a Unix Socket and relay the results back. This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.

At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results. These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling. The exception monad further ensures robust management of potential failures and unexpected events.

Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server. Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.

\section{\texttt{conn.c}}

\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. It is a part of the "King" (a.k.a. "Urth") process. It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.

The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane. It serves as a way for unix processes to receive insights about what is happening on urbit. It also allows for pass commands either \%khan and the injection of raw kernel moves into the system. The implication of this means that whoever is running your urbit runtime controls your urbit.

From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}. 
The newt-encoded format is:

\begin{verbatim}
V.BBBB.JJJJ.JJJJ...
\end{verbatim}

\begin{itemize}
  \item \texttt{V} is the version
  \item \texttt{B} is the jam size in bytes (little endian)
  \item \texttt{J} is the jammed noun (little endian)
\end{itemize}

This structure allows for a variety of commands to be executed, including

\begin{enumerate}
  \item \texttt{\%ovum} - the injection of raw kernel moves 
  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
  \item \texttt{\%peek} - namespace scry requests into Arvo
  \item \texttt{\%peel} - emulated namespace scry requests into Vere
\end{enumerate}

A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
\begin{itemize}
  \item \texttt{request-id} matches the input \texttt{request-id}
  \item \texttt{output} depends on the \texttt{command}
\end{itemize}

An invalid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[0 \%bail error-code error-string]}.

\section{\%lick - The Low Level IPC Interface}

Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.  (This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.

\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports. Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.  These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
The connexions are made via filepaths in \texttt{.urb/dev} of the pier.

The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.

To understand what \texttt{\%lick} is doing, we need to look at Unix's IPC model briefly.  IPC ("interprocess communication") describes any way that two processes in an operating system's shared context have to communicate with each other.  \texttt{\%lick} focuses on Unix domain sockets\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}}, which are just communication endpoints\footnote{\url{https://man7.org/linux/man-pages/man7/unix.7.html}}.

For instance, a valid use of \texttt{\%lick} would use cards that look like this:

\begin{verbatim}
++  init  [[%pass / %arvo %l %spin /control]~ this]
::
++  on-arvo
  |=  [=wire =sign-arvo]
  ?+  sign-arvo  (on-arvo:def wire sign-arvo)
      [%lick %soak *]
      ?+  mark.sign-arvo  [~ this]
      ::
        %connect
      ~&  >  "connect"
      :_  this  
          [%pass /spit %arvo %l %spit /control %init area.state]~
      ==  
  ==
::
++  send-state
  |=  =state
  ^-  card:agent:gall
  :*  %pass 
      /spit 
      %arvo %l 
      %spit 
      /control 
      %state 
      [slick:state face.state food.state live.state]
   ==
\end{verbatim}

The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.  The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.

Arvo will send three types of \texttt{\%soak}s to an open \texttt{\%lick} port. These \texttt{\%soak}s are \texttt{\%connect} when the first connection is established on the Earth-side IPC port, \texttt{\%disconnect} when the last connection is broken on the Earth-side IPC port, and \texttt{\%error} when an error occurs. \texttt{\%lick} also will send a \texttt{\%disconnect} \texttt{\%soak} to every agent when Vere is started.

Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.  See e.g. \texttt{\%++ap-generic-take}.  This lets multiple agents share sockets with the same name, and each agent can have its own folder.

@@

As Urbit had to deal with the reality of running on Earth software, Developers had to come up with various ways of communicating between these systems. Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic.

Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the original idea of communication between Earth and Mars. These vanes force Earth to speak to Mars in a language it can understand, nouns.

\printbibliography
\end{document}
